---
    output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# vlad 
<!-- [![Build Status](https://travis-ci.org/wittenberg/vlad.svg)](https://travis-ci.org/wittenberg/vlad)  -->
<!-- [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/vlad)](http://cran.r-project.org/package=vlad)  -->
<!-- [![Coverage Status](https://codecov.io/gh/wittenberg/vlad/graph/badge.svg)](https://codecov.io/github/wittenberg/vlad?branch=master)  -->
<!-- [![Downloads](https://cranlogs.r-pkg.org/badges/vlad)](https://CRAN.R-project.org/package=vlad) -->
<!-- [![Total Downloads](https://cranlogs.r-pkg.org/badges/grand-total/vlad?color=orange)](https://CRAN.R-project.org/package=vlad) -->

An R-package which contains functions to set up risk-adjusted quality control charts in health care.

## Main features

- Risk-adjusted CUSUM chart based on E-O by Wittenberg et al. (2018)
- Risk-adjusted CUSUM chart based on log-likelihood ratio statistic by Steiner et al. (2000)
- Algorithms are implemented using Rcpp and RcppArmadillo
- High performance with parallel computation

## Installation

You can install the released version of **vlad** from [CRAN](https://cran.r-project.org/web/packages/vlad/index.html) with:

``` r
install.packages("vlad")
```

And the development version from [GitHub](https://github.com/wittenberg/vlad) with:

``` r
# install.packages("devtools")
devtools::install_github("wittenberg/vlad")
```
## Example


Load libraries:
```{r load libraries, message=FALSE}
library("vlad")
library("dplyr")
library("tidyr")
library("ggplot2")
```

Subset the dataset `cardiacsurgery` into Phase I (first two years) and Phase II (five years) and estimate a risk model based on `phaseI`.
```{r estimate risk model}
data("cardiacsurgery", package = "spcadjust")
cardiacsurgery <- cardiacsurgery %>% rename(s = Parsonnet) %>%
  mutate(y = ifelse(status == 1 & time <= 30, 1, 0),
        phase = factor(ifelse(date < 2*365, "I", "II")))
head(cardiacsurgery)
phaseI <- filter(cardiacsurgery, phase == "I") %>% select(s, y)
coeff <- round(coef(glm(y ~ s, data = phaseI, family = "binomial")), 3)
print(coeff)
```

### Create VLADs for seven surgeons
By using the estimated risk model coefficients `coeff`, for each pair of Parsonnet score `s` and operation outcome values `y`, the difference between expected and observed outcome is calculated with the function `calceo()`.
Thereafter, differences are cummulated to create the VLAD. This is done for all seven surgeons of the `cardiacsurgery` dataset. Results are saved to the object `vlads7`.
```{r vlads7}
vlads7 <- lapply(1:7, function(j){
  Si <- filter(cardiacsurgery, surgeon == j)
  EO <- sapply(seq_along(Si$s), function(i) calceo(df = Si[i, c("s", "y")], coeff = coeff))
  select(Si, surgeon, phase) %>%  mutate(n = 1:length(EO), cEO = cumsum(EO))
}) 

```

Create Variable life-adjusted Displays for each surgeon from the object `vlads7`.
```{r VLADS1-7, fig.align='center', fig.width=8, fig.height=10}
vlads7 %>% 
  bind_rows() %>%  
  gather(key = "Surgeon", value = value, c(-n, -surgeon, -phase)) %>%
  ggplot(aes(x = n, y = value, colour = phase, group = Surgeon)) +
    geom_hline(yintercept = 0, colour = "darkgreen", linetype = "dashed") +
    geom_line(size = 1.1) + facet_wrap( ~ surgeon, ncol = 2, scales = "free") +
    labs(x="Patient number n", y="CUSUM E-O") + theme_classic() +
    scale_y_continuous(sec.axis = dup_axis(name = NULL, labels = NULL)) +
    scale_x_continuous(sec.axis = dup_axis(name = NULL, labels = NULL))
```

### Create a VLAD for surgeon 2
```{r vladS2, fig.align='center'}
S2 <- filter(cardiacsurgery, surgeon == 2) %>% select(phase, s, y)
S2I <- subset(S2, c(phase == "I"))
S2II <- subset(S2, c(phase == "II"))
coeff <- coef(glm(y ~ s, data = S2I, family = "binomial"))
EO <- sapply(1:nrow(S2), function(i) calceo(df = S2[i, c("s", "y")], coeff = coeff))

df1 <- select(S2, phase) %>% mutate(n = row_number(), cEO = cumsum(EO))
df2 <- gather(df1, variable, value, c(-n, -phase))

p1 <- ggplot(df2, aes(x = n, y = value, colour = phase)) +
  geom_hline(yintercept = 0, linetype = "dashed") + geom_line() + geom_point() + 
  labs(x = "Patient number", y = "CUSUM E-O") + theme_classic() +
  scale_y_continuous(sec.axis = dup_axis(name = NULL, labels = NULL)) +
  scale_x_continuous(sec.axis = dup_axis(name = NULL, labels = NULL))
p1
```

### Compute thresholds of a risk-adjusted CUSUM chart for surgeon 2 

 
Upper and lower control limits of the risk-adjusted CUSUM chart based on log-likelihood ratio statistic can be computed with the function `racusum_arl_h_sim()`. The implemention uses parallel simulation and a multi-stage search procedure.  
```{r}
# set a random number generator for parallel computations
RNGkind("L'Ecuyer-CMRG")
# number of simulation runs
m <- 10^4
# assign cores
nc <- parallel::detectCores()
# verbose calculation 
UCL_sim <- racusum_crit_sim(L0 = 740, df = S2I[, c("s", "y")], coeff = coeff, m = m, RA = 2, nc = nc, verbose = TRUE)
# quite calculation
LCL_sim <- racusum_crit_sim(L0 = 740, df = S2I[, c("s", "y")], coeff = coeff, m = m, RA = 1/2, nc = nc, verbose = FALSE)
round(cbind(UCL_sim, LCL_sim), 3)
```

### References

Wittenberg et al. (2018). [A simple signaling rule for variable life‐adjusted display derived from an equivalent risk‐adjusted CUSUM chart](https://doi.org/10.1002/sim.7647)

Steiner et al. (2000). [Monitoring surgical performance using risk-adjusted cumulative sum charts](https://doi.org/10.1093/biostatistics/1.4.441) 

### Authors

Philipp Wittenberg and Sven Knoth

### License

GPL (>= 2)
