---
title: "Vignette for the *vlad* package"
author: "by [Philipp Wittenberg](https://github.com/wittenberg/) - `r format(Sys.time(), '%d %B %Y')`"
bibliography: VLAD.bib
pandoc_args: ["--natbib"]
biblio-style: plain
link-citations: true
header-includes:
   - \usepackage[utf8]{inputenc}
   - \usepackage[T1]{fontenc}
   - \usepackage{moreverb}
   - \usepackage{amsmath}
   - \usepackage{textcomp}
   - \usepackage{graphicx}
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Construct CUSUM charts for hospital performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, cache=FALSE, fig.align="center", comment="#>", eval=TRUE)
```
\[
\DeclareMathOperator*{\gleich}{=}
\]


Install the *vlad* package from CRAN

```{r}
library("vlad")
```

He says @Wittenberg.etal_2018.
[@Wittenberg.etal_2018]

```{r, eval=TRUE}
data("cardiacsurgery", package="spcadjust") # example data set
library("parallel")                         # parallel simulations
library("dplyr")                            # data handling/manipulation
library("tidyr")                            # data preparation for plotting
library("ggplot2")                          # for plotting

theme_set(theme_bw() +
            theme(panel.grid.major=element_blank(),
                  panel.grid.minor=element_blank(),
                  plot.title=element_text(hjust=.5))
          )


SALL <- cardiacsurgery %>% rename(s = Parsonnet) %>%
  mutate(y = ifelse(status == 1 & time <= 30, 1, 0),
         phase = factor(ifelse(date < 2*365, "I", "II")))

SI   <- subset(SALL, c(phase == "I"))
GLM  <- glm(y ~ s, data=SI, family="binomial")
pi1  <- fitted.values(GLM)
pmix <- data.frame(SI$y, pi1, pi1)

L0 <- 740
m  <- 1e2
nc <- detectCores()

k1 <- optimal_k(pmix=pmix, RA=1/2)
k2 <- optimal_k(pmix=pmix, RA=2)
h1 <- eocusum_crit_sim(L0=L0, pmix=pmix, k=k1, side="up",  m=m, nc=nc)
h2 <- eocusum_crit_sim(L0=L0, pmix=pmix, k=k2, side="low", m=m, nc=nc)

## Detecting improvement: k1, h1, d1, theta1
## Detecting deterioration: k2, h2, d2, theta2
d1 <- h1 / k1
d2 <- h2 / k2
theta1 <- atan(k1)*180 / pi
theta2 <- atan(k2)*180 / pi
p <- data.frame(h1, h2, k1, k2, d1, d2, theta1, theta2)
p

S2II <- subset(SALL, c(phase == "II" & surgeon == 2))
pi2  <- predict(GLM, type="response", newdata=data.frame(s=S2II$s))

z  <- pi2-S2II$y
Sn <- cumsum(z)
VM <- compute_vmask(z, p$d1, p$d2, p$theta1, p$theta2)
data.frame(n=1:length(Sn), value=Sn) %>%
  ggplot(aes(x=n, y=value)) +
  geom_hline(yintercept=0, colour="darkgreen", linetype="dashed") +
  ## VLAD
  geom_line() +
  ## V-mask arms
  geom_line(data=VM$arms, aes(group=interaction(group, number), colour=Masks), alpha=.3) +
  ## V-mask vertical nose
  geom_line(data=VM$nose, aes(group=interaction(group, number), colour=Masks), alpha=.3) +
  ## alarm points
  geom_point(data=VM$alarms, aes(colour=Masks), size=2) +
  labs(x="Patient number n", y="E-O CUSUM", title=paste0("Surgeon: ", 2)) +
  guides(colour = "none")
```

## Control charts performance evaluation
Compute out of Control ARLs of single sided control charts RA-CUSUM, EO-CUSUM with $k=kopt$ and $k=0$ for detecting deterioration.
```{r, results=FALSE, eval=FALSE}
L0 <- 7500                                             # In control ARL of single sided control chart
m <- 1e4                                                # Increase simulation size for higher accurracy
nc <- detectCores()                                     # Number of cores used for parallel simulations 
RQv <- unique(c(seq(1, 2, .05), seq(2, 4, .1)))         # define Out-of-control ARL states
GLM <- glm(y ~ s, data=SALL, family="binomial")         # use the complete data set
pi1 <- fitted.values(GLM)                               # Predicted outcome from the risk-model
pmix <- data.frame(SALL$y, pi1, pi1)                    # build paient mix
# calibrate all three control charts to Incontrol ARL (L0)
h1 <- racusum_crit_sim(L0, pmix=pmix, RA=2, m=m, nc=nc)
k2 <- optimal_k(pmix=pmix, RA=2)
h2 <- eocusum_crit_sim(L0=L0, pmix=pmix, k=k2, side="low", m=m, nc=nc)
h3 <- eocusum_crit_sim(L0=L0, pmix=pmix, k=0,  side="low", m=m, nc=nc)

cl <- makeCluster(getOption("cl.cores", nc))            # set up a cluster for parallel simulation
clusterExport(cl, c("h1", "k2", "h2", "h3", "pmix", "m",# export variables and functions for parallel processing
                    "RQv", "eocusum_arl_sim","racusum_arl_sim"))
# Empty Run lengths array
mylist <- list(1:3, RQv, 1:m)
names(mylist) <- c("chart", "RQ", "m")
RLS <- array(NA, dim=c(3, length(RQv), m), dimnames=mylist)
## Parallel simulation
for (RQ in RQv) {
  # RA CUSUM
  RLS[1, which(RQv==RQ), ] <- parSapply(cl, 1:m, racusum_arl_sim, pmix=pmix, h=h1, RQ=RQ, RA=2)
  # EO CUSUM with k=kopt
  RLS[2, which(RQv==RQ), ] <- parSapply(cl, 1:m, eocusum_arl_sim, pmix=pmix, k=k2, h=h2, RQ=RQ, side="low")
  # EO CUSUM with k=0
  RLS[3, which(RQv==RQ), ] <- parSapply(cl, 1:m, eocusum_arl_sim, pmix=pmix, k=0,  h=h3, RQ=RQ, side="low")
  cat("RQ: ", RQ, "\n")
}
stopCluster(cl)                                         # close cluster for parallel simulation
ARL   <- apply(RLS, c(1, 2), mean)
ARLSE <- apply(RLS, c(1, 2), sd)/sqrt(m)
dat   <- data.frame(cbind(RQv, 
                        "ARL_LLR" =ARL[1, ], "SE_LLR" =ARLSE[1, ], 
                        "ARL_kopt"=ARL[2, ], "SE_kopt"=ARLSE[2, ], 
                        "ARL_k0"  =ARL[3, ], "SE_k0"  =ARLSE[3, ]))
```

```{r, eval=FALSE}
# show results for selected out of contol states
filter(dat, RQv %in% c(1, 1.05, 2, 4)) %>% mutate_if(is.numeric, round, 2)
```

Compute out of Control ARLs of single sided control charts RA-CUSUM, EO-CUSUM with $k=kopt$ and $k=0$ for detecting improvement.
```{r, results=FALSE, eval=FALSE}
L0 <- 7500                                              # In control ARL of single sided control chart
m <- 1e4                                                # Increase simulation size for higher accurracy
nc <- detectCores()                                     # Number of cores used for parallel simulations 
RQv <- 1/unique(c(seq(1, 2, .05), seq(2, 4, .1)))       # define Out-of-control ARL states
GLM <- glm(y ~ s, data=SALL, family="binomial")         # use the complete data set
pi1 <- fitted.values(GLM)                               # Predicted outcome from the risk-model
pmix <- data.frame(SALL$y, pi1, pi1)                    # build paient mix
# calibrate all three control charts to Incontrol ARL (L0)
h1 <- racusum_crit_sim(L0, pmix=pmix, RA=1/2, m=m, nc=nc)
k1 <- optimal_k(pmix=pmix, RA=1/2)
h2 <- eocusum_crit_sim(L0=L0, pmix=pmix, k=k1, side="up", m=m, nc=nc)
h3 <- eocusum_crit_sim(L0=L0, pmix=pmix, k=0,  side="up", m=m, nc=nc)

cl <- makeCluster(getOption("cl.cores", nc))            # set up a cluster for parallel simulation
clusterExport(cl, c("h1", "k1", "h2", "h3", "pmix", "m",# export variables and functions for parallel processing
                    "RQv", "eocusum_arl_sim","racusum_arl_sim"))
# Empty Run lengths array
mylist <- list(1:3, RQv, 1:m)
names(mylist) <- c("chart", "RQ", "m")
RLS <- array(NA, dim=c(3, length(RQv), m), dimnames=mylist)
## Parallel simulation
for (RQ in RQv) {
  # RA CUSUM
  RLS[1, which(RQv==RQ), ] <- parSapply(cl, 1:m, racusum_arl_sim, pmix=pmix, h=h1, RQ=RQ, RA=1/2)
  # EO CUSUM with k=kopt
  RLS[2, which(RQv==RQ), ] <- parSapply(cl, 1:m, eocusum_arl_sim, pmix=pmix, k=k1, h=h2, RQ=RQ, side="up")
  # EO CUSUM with k=0
  RLS[3, which(RQv==RQ), ] <- parSapply(cl, 1:m, eocusum_arl_sim, pmix=pmix, k=0,  h=h3, RQ=RQ, side="up")
  cat("RQ: ", RQ, "\n")
}
stopCluster(cl)                                         # close cluster for parallel simulation
ARL   <- apply(RLS, c(1, 2), mean)
ARLSE <- apply(RLS, c(1, 2), sd)/sqrt(m)
dat2  <- data.frame(cbind(RQv, 
                        "ARL_LLR" =ARL[1, ], "SE_LLR" =ARLSE[1, ], 
                        "ARL_kopt"=ARL[2, ], "SE_kopt"=ARLSE[2, ], 
                        "ARL_k0"  =ARL[3, ], "SE_k0"  =ARLSE[3, ]))
```

```{r, eval=FALSE}
# show results for selected out of contol states
filter(dat2, RQv %in% c(1, 1/1.05, 1/2, 1/4)) %>% mutate_if(is.numeric, round, 2)
```

Plot results.
```{r, fig.width=10, eval=FALSE}
bind_rows(
  select(dat,  RQv, ARL_LLR, ARL_kopt, ARL_k0),
  select(dat2, RQv, ARL_LLR, ARL_kopt, ARL_k0),
  .id="Case") %>% 
  rename("RQ"=RQv, "LLR CUSUM"=ARL_LLR, "EO CUSUM, k=0"=ARL_k0, "EO CUSUM, k=kopt"=ARL_kopt) %>%
  mutate(Case=fct_recode(Case, "Deterioration: RA=2"="1", "Improvement: RA=1/2"="2")) %>%
  gather("Chart", ARL, c(-RQ, -Case)) %>%
  ggplot(aes(x=RQ, y=ARL, group=Chart, colour=Chart)) +
    geom_line() +
    scale_y_log10() +
    facet_wrap(~Case, scales="free")
```


Philipp Wittenberg.

### License

GPL (>= 2)

## References
